## Unsupervised Machine Learning - Test Your Knowledge Solution

```{r, results=FALSE, message=FALSE}
library(ggplot2)
library(pheatmap)
library(tidyverse)
library(janitor)
library(factoextra)
#library(tibble)
```

#### Load and Organize Data for Pesticides and Flame Retardants (FR)
```{r}
# Pesticide and Flame Retardant data 
chems <- read.csv("Module5_4_TYKInput.csv", fileEncoding = "UTF-8-BOM")
chems <- chems %>% column_to_rownames("CASRN")
colnames(chems)
head(chems)

# Isolate just the physchem properties
props <- chems %>% select(!List)
colnames(props)
head(props)
```

#### Plot original chemical properties as first look at data
```{r}
ggplot(chems, aes(x = ATMOSPHERIC_HYDROXYLATION_RATE_.AOH._CM3.MOLECULE.SEC_OPERA_PRED, y = OCTANOL_WATER_PARTITION_LOGP_OPERA_PRED, color = chems$List)) +
  geom_point(size = 2) + theme_bw() + 
  ggtitle('Bivariate Plot of Two Original Physchem Variables') + 
  xlab("Atmospheric Hydroxylation Rate") + ylab("Octanol Water Partition Coefficient")

```

Look at another set of properties
```{r}
ggplot(chems, aes(x = SOIL_ADSORPTION_COEFFICIENT_KOC_L.KG_OPERA_PRED, y = BOILING_POINT_DEGC_OPERA_PRED, color = chems$List)) + 
  geom_point(size = 2) + theme_bw() + 
  ggtitle('Bivariate Plot of Two Original Physchem Variables') + 
  xlab("Soil Adsorption Coefficient") + ylab("Boiling Point")

```

There are no clear associations when only looking at two variables at a time, so let's proceed with running *k*-means.

### Question #1: What are some of the physicochemical properties that seem to be driving chemical clustering patterns derived through *k*-means?

Specify we are interested in 2 clusters
```{r}
num.centers <- 2
```

Run clustering analysis
```{r}
clusters <- kmeans(props,        # input dataframe
                   centers = num.centers,  # number of cluster centers to calculate
                   iter.max = 1000,        # the maximum number of iterations allowed
                   nstart = 50)            # number of rows used as the random set for the initial centers
```

Extract cluster assignments for the chemicals
```{r}
clusters$cluster
```

Aggregate results and prepare dataset for making a heatmap
```{r}
dat_wclusters <- as.data.frame(cbind(props,clusters$cluster))
colnames(dat_wclusters)[13] <- "kmeans_cluster"  # renaming this new column "kmeans_cluster"
dat_wclusters <- dat_wclusters[order(dat_wclusters$kmeans_cluster),]  # sorting data by cluster assignments
colnames(dat_wclusters)


hm_cluster <- data.frame(dat_wclusters$kmeans_cluster, row.names=row.names(dat_wclusters))  # creating the dataframe to show which chemicals are assigned to which cluster for heatmap annotation bar
colnames(hm_cluster) <- "kmeans_cluster"   # reassigning the column name
hm_cluster$kmeans_cluster <- as.factor(hm_cluster$kmeans_cluster)   # coercing the cluster numbers into factor variables, to make the heat map prettier
head(hm_cluster)
```
<br>

Make heatmap
```{r, fig.height=8, fig.width=10}
chem_hm <- pheatmap(dat_wclusters[,1:10], 
                    main = "Heatmap of Physicochemical Properties with k-means Cluster Assignments",
                    cluster_rows = FALSE, cluster_cols = FALSE, 
                    scale = "column",               
                    annotation_row = hm_cluster,  
                    angle_col = 45, fontsize_col = 5, fontsize_row = 3,  
                    cellheight = 3, cellwidth = 25,  
                    border_color = FALSE  
)
```
**Question #1**: What are some of the physicochemical properties that seem to be driving chemical clustering patterns derived through k-means?

**Answer**: Properties with values that show obvious differences between resulting clusters including atmospheric hydroxylation, bioconcentration factor, boiling point, and soil adsorption coefficient.

<br>

### Question 2: Upon reducing the data through PCA, which physicochemical property contributes the most towards informing data variance captured in the primary principal component (Comp.1)?

Run PCA
```{r}
colnames(props) # review dataset of just physicochemical property data
my.pca <- prcomp(props,   # input dataframe of physchem data
                   scale = TRUE, center = TRUE) # scaling and centering data
```
Looking at how much of the variance was able to be compressed into the eigenvectors
```{r}
fviz_eig(my.pca, addlabels = TRUE)
```

The scree plot shows that the first two dimensions (PCs) capture less than 50% of the data, which isn't ideal but we'll proceed forward.

View scores for each chemical's contribution towards each principal component
```{r}
head(my.pca$x)
```
View loadings
```{r}
my.pca$rotation
```
**Question 2**: Upon reducing the data through PCA, which physicochemical property contributes the most towards informing data variance captured in the primary principal component (Comp.1)?

**Answer**: Boiling point contributes the most towards principal component #1, as it has the largest magnitude (-0.436).

<br>

### Question 3: Construct a PCA plot of the first two Principal Components for this dataset. MAKE A BETTER QUESTION

**Answer**:
```{r}
ggplot(as.data.frame(my.pca$x), aes(x = PC1, y = PC2, color = as.factor(chems$List))) + 
  geom_point(size = 2) +
  theme_bw() + 
  ggtitle('PCA Plot of the First 2 PCs, colored by Chemical Class') + 
  xlab("Principal Component 1 (33.8%)") + ylab("Principal Component 2 (14.9%)") 
```
